Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-26 reanalysis-1-data_preparation.qmd
refactor for new requirements

outcome_vars <- c(
  "outcome_VA_logMAR",
  "person_ever_lapse"
)

@@

1. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)

- Clinical assumption: These conditions would be documented if present
- This recoding happens BEFORE propensity score modeling
- Maintains consistency with original analysis,

2. **CCI/DCSI**:

@@

3. **New analysis variables**:

@@

```{r create-variables}
# Step 1: Create person_dr first (needed for imputation)
df_analysis <- df_combined |>
  create_person_dr()  # Creates person_dr from No_DR/NPDR/PDR

# Step 2: Create other core variables
df_analysis <- df_analysis |>
  mutate(
    # Exposure variable
    ever_lapse_binary = if_else(person_ever_lapse == TRUE, 1, 0),

@@

# Step 3: Create derived variables using transformation functions
df_analysis <- df_analysis |>
  create_outcome_variables() |>  # Creates outcome_va_vi_binary from logMAR
  create_age_cat() |>  # Creates age_cat from continuous age
  create_treatment_type() |>  # Creates treatment_type and any_treatment
  create_dr_severity_variables()  # Creates person_dr_severity and dr_severity factor
```

@@

# Assess missing data AFTER recoding
missing_summary <- df_analysis |>
  dplyr::select(dplyr::all_of(c(ps_model_vars, "outcome_va_vi_binary"))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  dplyr::mutate(
    pct_missing = round(n_missing / nrow(df_analysis) * 100, 2),
    var_type = dplyr::case_when(
      variable %in% c("outcome_va_vi_binary") ~ "Outcome",
      variable == "ever_lapse_binary" ~ "Exposure",
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Covariate (recoded)",
      TRUE ~ "Covariate"
    ),
    handling = dplyr::case_when(
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Missing → 0 (no condition)",
      variable %in% c("CCI", "DCSI") ~ "To be imputed",
      TRUE ~ "To be imputed"
    )
  ) |>
  dplyr::arrange(desc(n_missing))

@@

@@

variable_lists <- list(
  ps_model_vars = ps_model_vars,
  outcome_model_vars = outcome_model_vars,
  dr_vars = dr_vars,
  all_analysis_vars = all_analysis_vars
)
saveRDS(variable_lists, file.path(reanalysis_data_dir, "variable_lists.rds"))

message("Data saved to: ", reanalysis_data_dir)
```

>>>>

outcome_vars <- c(
  "outcome_VA_logMAR",
  "person_ever_lapse",
  "cohort_id"  # Indicates if true 2-year outcome is available
)

@@

1. **Visual Acuity Outcomes (Temporal Structure)**:

- `cohort_id = "submitted"`: Has true 2-year outcome VA
- `cohort_id = "excluded"`: Lacks true 2-year outcome
  - Any existing outcome_VA_logMAR value is an intermediate follow-up
  - These intermediate values will be moved to a new variable
  - True 2-year outcome will be imputed using temporal structure

2. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)

- Clinical assumption: These conditions would be documented if present
- This recoding happens BEFORE propensity score modeling
- Maintains consistency with original analysis

3. **CCI/DCSI**:

@@

4. **New analysis variables**:

@@

```{r create-variables}
# Step 1: Handle temporal VA structure
# Create intermediate_VA_logMAR and adjust outcome_VA_logMAR based on cohort_id
df_analysis <- df_combined |>
  mutate(
    # Initialize intermediate VA variable
    intermediate_VA_logMAR = NA_real_,

    # For "excluded" cohort, move outcome values to intermediate
    # These are intermediate follow-ups, not true 2-year outcomes
    intermediate_VA_logMAR = if_else(
      cohort_id == "excluded" & !is.na(outcome_VA_logMAR),
      outcome_VA_logMAR,
      intermediate_VA_logMAR
    ),

    # Set outcome to NA for "excluded" cohort (will be imputed)
    outcome_VA_logMAR = if_else(
      cohort_id == "excluded",
      NA_real_,
      outcome_VA_logMAR
    ),

    # Convert cohort_id to factor
    cohort_id = factor(cohort_id)
  )

# Step 2: Create person_dr (needed for imputation)
df_analysis <- df_analysis |>
  create_person_dr()  # Creates person_dr from No_DR/NPDR/PDR

# Step 3: Create other core variables
df_analysis <- df_analysis |>
  mutate(
    # Exposure variable
    ever_lapse_binary = if_else(person_ever_lapse == TRUE, 1, 0),

@@

# Step 4: Create derived variables using transformation functions
df_analysis <- df_analysis |>
  create_outcome_variables() |>  # Creates outcome_va_vi_binary from logMAR
  create_age_cat() |>  # Creates age_cat from continuous age
  create_treatment_type() |>  # Creates treatment_type and any_treatment
  create_dr_severity_variables()  # Creates person_dr_severity and dr_severity factor
```

@@

# Assess missing data AFTER recoding
missing_summary <- df_analysis |>
  dplyr::select(dplyr::all_of(c(ps_model_vars, "outcome_va_vi_binary", 
                                "outcome_VA_logMAR", "intermediate_VA_logMAR", "cohort_id"))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  dplyr::mutate(
    pct_missing = round(n_missing / nrow(df_analysis) * 100, 2),
    var_type = dplyr::case_when(
      variable %in% c("outcome_va_vi_binary", "outcome_VA_logMAR") ~ "Outcome",
      variable == "intermediate_VA_logMAR" ~ "Intermediate (temporal)",
      variable == "cohort_id" ~ "Cohort indicator",
      variable == "ever_lapse_binary" ~ "Exposure",
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Covariate (recoded)",
      TRUE ~ "Covariate"
    ),
    handling = dplyr::case_when(
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Missing → 0 (no condition)",
      variable == "cohort_id" ~ "Never missing (complete)",
      variable == "intermediate_VA_logMAR" ~ "To be imputed (temporal predictor)",
      variable == "outcome_VA_logMAR" ~ "To be imputed (respecting temporal order)",
      variable %in% c("CCI", "DCSI") ~ "To be imputed",
      TRUE ~ "To be imputed"
    )
  ) |>
  dplyr::arrange(desc(n_missing))

@@

## Temporal VA Structure Summary

```{r temporal-va-summary}
# Summary of temporal VA structure
temporal_summary <- df_analysis |>
  group_by(cohort_id) |>
  summarise(
    n = n(),
    n_has_outcome = sum(!is.na(outcome_VA_logMAR)),
    n_has_intermediate = sum(!is.na(intermediate_VA_logMAR)),
    n_has_baseline = sum(!is.na(baseline_VA_logMAR)),
    .groups = "drop"
  ) |>
  mutate(
    pct_has_outcome = round(100 * n_has_outcome / n, 1),
    pct_has_intermediate = round(100 * n_has_intermediate / n, 1),
    pct_has_baseline = round(100 * n_has_baseline / n, 1)
  )

kable(temporal_summary,
      caption = "Visual acuity data availability by cohort",
      col.names = c("Cohort", "N", "Has Outcome", "Has Intermediate", "Has Baseline",
                   "% Outcome", "% Intermediate", "% Baseline"))

# Verify the temporal recoding worked correctly
verification_results <- data.frame(
  Check = c("Excluded cohort with outcome VA (should be 0)",
            "Submitted cohort with intermediate VA (should be 0)"),
  Count = c(
    sum(df_analysis$cohort_id == "excluded" & !is.na(df_analysis$outcome_VA_logMAR)),
    sum(df_analysis$cohort_id == "submitted" & !is.na(df_analysis$intermediate_VA_logMAR))
  ),
  Status = c(
    ifelse(sum(df_analysis$cohort_id == "excluded" & !is.na(df_analysis$outcome_VA_logMAR)) == 0, "✓ Pass", "✗ Fail"),
    ifelse(sum(df_analysis$cohort_id == "submitted" & !is.na(df_analysis$intermediate_VA_logMAR)) == 0, "✓ Pass", "✗ Fail")
  )
)

kable(verification_results, caption = "Temporal recoding verification")
```

@@

variable_lists <- list(
  ps_model_vars = ps_model_vars,
  outcome_model_vars = outcome_model_vars,
  dr_vars = dr_vars,
  all_analysis_vars = all_analysis_vars,
  temporal_vars = c("cohort_id", "intermediate_VA_logMAR", "outcome_VA_logMAR", "baseline_VA_logMAR")
)
saveRDS(variable_lists, file.path(reanalysis_data_dir, "variable_lists.rds"))

# Create a variable to show in text
data_save_path <- reanalysis_data_dir
```

**Data saved to:** `r data_save_path`

### 2025-08-26 reanalysis-2-multiple_imputation.qmd
refactor for new requirements

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 5 iterations (fast)
# final = 30 imputations, 20 iterations (slow but thorough)
analysis_mode <- "test"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 10
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

  # Outcome variable - SOURCE ONLY
  "outcome_VA_logMAR",  # Continuous version (source)
  # NOT outcome_va_vi_binary - derived from logMAR >= 0.3

@@

  # Demographics/Clinical
  "baseline_VA_logMAR",
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",  # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

@@

custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["age"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"

@@

# Ensure outcome is used to predict missingness (important for MAR assumption)
pred_matrix[, "outcome_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Don't predict itself

# Ensure exposure is used as predictor
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

@@

**Variables that will ACTUALLY be imputed** (have missing data, `r length(vars_to_impute)` total):
```{r vars-to-impute}
if (length(vars_to_impute) > 0) {
  data.frame(
    Variable = vars_to_impute,
    Missing_Count = final_methods |> 
      filter(Variable %in% vars_to_impute) |> 
      pull(N_Missing)
  ) |>
  kable()
} else {
  cat("No variables require imputation (all data complete)")
}

@@

**Complete variables with imputation methods assigned** (`r nrow(complete_with_methods)` total):
These variables have no missing data and won't be imputed, but have methods specified to ensure proper measurement level:
```{r complete-vars-table}
if (nrow(complete_with_methods) > 0) {
  kable(complete_with_methods, col.names = c("Variable", "Method Type"))
} else {
  cat("No complete variables with methods assigned")
}

@@

@@

```{r imputation-params}
imp_params <- data.frame(
  Parameter = c("Mode", "Number of imputations (m)", "Max iterations (maxit)", "Random seed"),
  Value = c(analysis_mode, n_imputations, n_iterations, seed)
)
kable(imp_params)
```

@@

mice_results <- mice(
  df_mice,
  m = n_imputations,
  method = custom_methods,
  predictorMatrix = pred_matrix,
  maxit = n_iterations,  # Uses n_iterations from setup
  seed = 2025,
  printFlag = TRUE
)

@@

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  cat("### Charlson Comorbidity Index (CCI)\n")
  cat("**Density plot:** Blue = observed data, Red = imputed values\n")
  densityplot(mice_results, ~ CCI) |> print()

  cat("\n**Stripplot:** Column 1 = original data (with gaps for missing), Columns 2+ = imputed datasets\n")
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
  cat("\n")
}

# DCSI - compare observed vs imputed  
if(sum(is.na(df_mice$DCSI)) > 0) {
  cat("### Diabetes Complications Severity Index (DCSI)\n")
  cat("**Density plot:** Blue = observed data, Red = imputed values\n")
  densityplot(mice_results, ~ DCSI) |> print()

  cat("\n**Stripplot:** Column 1 = original data (with gaps for missing), Columns 2+ = imputed datasets\n")
  stripplot(mice_results, DCSI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "DCSI Distribution: Original vs Imputed") |> print()
  cat("\n")
}

# Baseline VA - compare observed vs imputed
if(sum(is.na(df_mice$baseline_VA_logMAR)) > 0) {
  cat("### Baseline Visual Acuity (logMAR)\n")
  cat("**Density plot:** Blue = observed data, Red = imputed values\n")
  densityplot(mice_results, ~ baseline_VA_logMAR) |> print()

  cat("\n**Stripplot:** Column 1 = original data (with gaps for missing), Columns 2+ = imputed datasets\n")
  stripplot(mice_results, baseline_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Baseline VA Distribution: Original vs Imputed") |> print()
  cat("\n")
}

@@

>>>>

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 2 iterations (fast)
# final = 50 imputations, 20 iterations (slow but thorough)
analysis_mode <- "final"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 50
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

  # Temporal VA variables
  "cohort_id",  # Indicates if true 2-year outcome available
  "baseline_VA_logMAR",  # Baseline VA
  "intermediate_VA_logMAR",  # Intermediate follow-up VA (for excluded cohort)
  "outcome_VA_logMAR",  # 2-year outcome (true for submitted, missing for excluded)
  # NOT outcome_va_vi_binary - derived from logMAR >= 0.3

@@

  # Demographics/Clinical
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",  # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

@@

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),  # Ensure cohort_id is a factor
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  )
```

@@

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"  # Impute if missing
custom_methods["intermediate_VA_logMAR"] <- "pmm"  # IMPUTE intermediate values
custom_methods["outcome_VA_logMAR"] <- "pmm"  # Impute 2-year outcomes
custom_methods["cohort_id"] <- ""  # Never impute - complete indicator

# Other continuous variables
custom_methods["CCI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["DCSI"] <- "pmm"  # Predictive mean matching for count data
custom_methods["age"] <- "pmm"

@@

# TEMPORAL STRUCTURE RULES
# 1. Baseline can predict intermediate and outcome
pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1

# 2. Intermediate can predict outcome
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

# 3. Cohort_id can predict VA variables (indicates data structure)
pred_matrix["baseline_VA_logMAR", "cohort_id"] <- 1
pred_matrix["intermediate_VA_logMAR", "cohort_id"] <- 1
pred_matrix["outcome_VA_logMAR", "cohort_id"] <- 1

# 4. CRITICAL: Prevent reverse-time prediction
pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict intermediate
pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0  # Outcome cannot predict baseline
pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0  # Intermediate cannot predict baseline

# Cohort_id should not be predicted by anything
pred_matrix["cohort_id", ] <- 0

# Don't use variables as their own predictors
diag(pred_matrix) <- 0

# Ensure exposure is used as predictor for outcomes
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

@@

**Variables that will ACTUALLY be imputed** (have missing data, `r length(vars_to_impute)` total):
```{r vars-to-impute}
if (length(vars_to_impute) > 0) {
  data.frame(
    Variable = vars_to_impute,
    Missing_Count = final_methods |> 
      filter(Variable %in% vars_to_impute) |> 
      pull(N_Missing)
  ) |>
  kable()
} else {
  no_imputation_msg <- "No variables require imputation (all data complete)"
  print(no_imputation_msg)
}

@@

**Complete variables with imputation methods assigned** (`r nrow(complete_with_methods)` total):
These variables have no missing data and won't be imputed, but have methods specified to ensure proper measurement level:
```{r complete-vars-table}
if (nrow(complete_with_methods) > 0) {
  kable(complete_with_methods, col.names = c("Variable", "Method Type"))
} else {
  no_complete_msg <- "No complete variables with methods assigned"
  print(no_complete_msg)
}

@@

### Visit Sequence

```{r visit-sequence}
# Define the order of imputation to respect temporal flow
# Impute earlier timepoints before later ones
visit_sequence <- c()

# First impute baseline if needed
if ("baseline_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "baseline_VA_logMAR")
}

# Then impute intermediate
if ("intermediate_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "intermediate_VA_logMAR")
}

# Then impute outcome
if ("outcome_VA_logMAR" %in% vars_to_impute) {
  visit_sequence <- c(visit_sequence, "outcome_VA_logMAR")
}

# Add other variables that need imputation
other_vars_to_impute <- vars_to_impute[!vars_to_impute %in% 
                                       c("baseline_VA_logMAR", 
                                         "intermediate_VA_logMAR", 
                                         "outcome_VA_logMAR")]
visit_sequence <- c(visit_sequence, other_vars_to_impute)

# Store for display
imputation_order <- paste(visit_sequence, collapse = ", ")
```

**Imputation order (respecting temporal structure):** `r imputation_order`

@@

```{r imputation-params}
imp_params <- data.frame(
  Parameter = c("Mode", "Number of imputations (m)", "Max iterations (maxit)", "Random seed", "Visit sequence"),
  Value = c(analysis_mode, n_imputations, n_iterations, seed, "Temporal order enforced")
)
kable(imp_params)
```

@@

mice_results <- mice(
  df_mice,
  m = n_imputations,
  method = custom_methods,
  predictorMatrix = pred_matrix,
  visitSequence = visit_sequence,  # Enforce temporal order
  maxit = n_iterations,  # Uses n_iterations from setup
  seed = 2025,
  printFlag = TRUE
)

@@

# CCI - compare observed vs imputed
if(sum(is.na(df_mice$CCI)) > 0) {
  # Density plot: Blue = observed data, Red = imputed values
  densityplot(mice_results, ~ CCI, 
              main = "Charlson Comorbidity Index (CCI)") |> print()

  # Stripplot: Column 1 = original data, Columns 2+ = imputed datasets
  stripplot(mice_results, CCI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "CCI Distribution: Original vs Imputed") |> print()
}

# DCSI - compare observed vs imputed  
if(sum(is.na(df_mice$DCSI)) > 0) {
  densityplot(mice_results, ~ DCSI,
              main = "Diabetes Complications Severity Index (DCSI)") |> print()

  stripplot(mice_results, DCSI ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "DCSI Distribution: Original vs Imputed") |> print()
}

# Baseline VA - compare observed vs imputed
if(sum(is.na(df_mice$baseline_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ baseline_VA_logMAR,
              main = "Baseline Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, baseline_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Baseline VA Distribution: Original vs Imputed") |> print()
}

# Intermediate VA - compare observed vs imputed
if(sum(is.na(df_mice$intermediate_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ intermediate_VA_logMAR,
              main = "Intermediate Visual Acuity (logMAR)") |> print()

  stripplot(mice_results, intermediate_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "Intermediate VA Distribution: Original vs Imputed") |> print()
}

# Outcome VA - compare observed vs imputed
if(sum(is.na(df_mice$outcome_VA_logMAR)) > 0) {
  densityplot(mice_results, ~ outcome_VA_logMAR,
              main = "Outcome Visual Acuity at 2 Years (logMAR)") |> print()

  stripplot(mice_results, outcome_VA_logMAR ~ .imp, pch = 20, cex = 1.2,
            xlab = "Dataset (1=Original, 2+=Imputations)",
            main = "2-Year Outcome VA Distribution: Original vs Imputed") |> print()
}

@@

### Temporal VA Imputation Summary

- **Baseline VA:** Starting point for temporal chain
- **Intermediate VA:** Bridge between baseline and 2-year outcome
- **Outcome VA:** Final 2-year outcome (missing for 'excluded' cohort)
- **Temporal order enforced:** baseline → intermediate → outcome
- **No reverse-time prediction allowed**

### 2025-08-26 reanalysis-4-outcome_analysis-interaction.qmd
refactor for new requirements

**DR Severity:**
```{r dr-severity-dist}
table(imputed_datasets[[1]]$dr_severity) |> kable()
```

**Treatment Status:**
```{r treatment-dist}
table(imputed_datasets[[1]]$any_treatment) |> kable()
```

**Lapse Status:**
```{r lapse-dist}
table(imputed_datasets[[1]]$ever_lapse_binary) |> kable()
```

@@

# Pivot to wide format for better readability
library(tidyr)
crosstab_wide <- crosstab_df |>
  pivot_wider(names_from = Lapse_Status, 
              values_from = Count,
              values_fill = 0) |>
  arrange(DR_Severity, Treatment)

kable(crosstab_wide, 
      caption = "Distribution of patients by DR severity, treatment, and lapse status",
      col.names = c("DR Severity", "Treatment", "No Lapse", "Lapsed"))
```

>>>>

**DR Severity:**
```{r dr-severity-dist}
dr_table <- table(imputed_datasets[[1]]$dr_severity)
dr_df <- data.frame(
  Category = names(dr_table),
  Count = as.numeric(dr_table),
  Percentage = paste0(round(100 * as.numeric(dr_table) / sum(dr_table), 1), "%")
)
kable(dr_df, align = c('l', 'r', 'r'))
```

**Treatment Status:**
```{r treatment-dist}
treatment_table <- table(imputed_datasets[[1]]$any_treatment)
treatment_df <- data.frame(
  Treatment = c("No Treatment", "Treatment"),
  Count = as.numeric(treatment_table),
  Percentage = paste0(round(100 * as.numeric(treatment_table) / sum(treatment_table), 1), "%")
)
kable(treatment_df, align = c('l', 'r', 'r'))
```

**Lapse Status:**
```{r lapse-dist}
lapse_table <- table(imputed_datasets[[1]]$ever_lapse_binary)
lapse_df <- data.frame(
  Lapse_Status = c("No Lapse", "Lapsed"),
  Count = as.numeric(lapse_table),
  Percentage = paste0(round(100 * as.numeric(lapse_table) / sum(lapse_table), 1), "%")
)
kable(lapse_df, align = c('l', 'r', 'r'))
```

@@

# Calculate total for percentages
total_n <- sum(crosstab_df$Count)

# Pivot to wide format for better readability with percentages
library(tidyr)
crosstab_wide <- crosstab_df |>
  pivot_wider(names_from = Lapse_Status, 
              values_from = Count,
              values_fill = 0) |>
  arrange(DR_Severity, Treatment) |>
  mutate(
    Total = `0` + `1`,
    No_Lapse_Pct = paste0(round(100 * `0` / Total, 1), "%"),
    Lapsed_Pct = paste0(round(100 * `1` / Total, 1), "%"),
    Row_Pct = paste0(round(100 * Total / total_n, 1), "%")
  ) |>
  select(DR_Severity, Treatment, 
         No_Lapse = `0`, No_Lapse_Pct,
         Lapsed = `1`, Lapsed_Pct,
         Total, Row_Pct)

kable(crosstab_wide, 
      caption = "Distribution of patients by DR severity, treatment, and lapse status",
      col.names = c("DR Severity", "Treatment", 
                    "No Lapse (n)", "No Lapse (%)",
                    "Lapsed (n)", "Lapsed (%)",
                    "Total (n)", "Row %"),
      align = c('l', 'l', 'r', 'r', 'r', 'r', 'r', 'r'))
```
